Fragile Families Data Cleaning

### 2021-11-11 Reports/Archive/try1.Rmd
moved some reports to archive folder

# Lab meeting preface & necessary background

**Purpose**: This presentation is a practice for presenting the preliminary results of the third aim of my dissertation to a committee member and important collaborator. I am seeking to get buy-in for splitting the analysis into two papers and begining to write the first paper.

**Objectives & desired feedback**: 
My objectives for this lab meeting are:

1. Practice presenting results 
2. Test drive presenting through Rmarkdown htmls

And correspondingly, I'm seeking feedback on the logical flow of the presentation and its understandability.

**Necessary background**:

#  Objectives and  hypotheses

A) Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen.

B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking.

C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and support vector machines and evaluate the agreement of the classifier at two time points.

Hypotheses:
- I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups. 
- I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time. 
- Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.

Additional

@@

## Labeling phenotype data
```{r, phenotypedataReadinLabel, include=F}
pheno<-readRDS(paste0(datadir, "OGData/fullpheno.rds"))

#Phenotype labeling & filtering to variables of interest
#labels
#to set labels using sjlabelled, needed to fix multibyte characters 
#https://stackoverflow.com/questions/4993837/r-invalid-multibyte-string
find_offending_character <- function(x, maxStringLength=256){  
  print(x)
  for (c in 1:maxStringLength){
    offendingChar <- substr(x,c,c)
    #print(offendingChar) #uncomment if you want the indiv characters printed
    #the next character is the offending multibyte Character
  }    
}
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\xad", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\x92", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
#define vars of interest 
idvars<-c("id", "ff_id")
demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")

#filter to these variables
#FF has data in original form +sjlabelled value labels
#since these dont' show in compareGroups
#also will make FF_labeled, where categorical, binary variables
#will be replaced with their labels
#FF<-pheno %>% select(one_of(idvars), one_of(demovars), one_of(smokingvars), one_of(prenataldrugusevar))
varIDX<-which(names(pheno)%in%c(idvars, demovars, smokingvars, prenataldrugusevar))
FF<-pheno[,varIDX]
varLabels<-attributes(pheno)$var.labels[varIDX]
varLabels[59]<-"ID"
FF<-set_label(FF, c(varLabels))
#create FF_labeled
#read in csv with metadata about variables 
FFmeta<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/FFMetadata_v07.csv")
#create long data dictionary  
FFlong<-FFmeta%>% select(old_name, contains('value'), contains('label')) %>% pivot_longer(cols = contains("value"), names_to="value", values_to="old_values")
FFlong<-FFlong%>%pivot_longer(cols = contains("label"), names_to="label", values_to="new_labels")
recodes_long<-FFlong %>% filter(gsub("value", "", value)==gsub("label", "", label)& !is.na(old_values)) %>% select(old_name, old_values, new_labels) 
#store continuous, binary, categorical etc
FFmyMeta<-FFmeta %>% filter(old_name %in% c(idvars, demovars, smokingvars, prenataldrugusevar))
myContinuous<-FFmyMeta %>% filter(type=="Continuous") %>%pull(old_name);
myContinuous<-colnames(FF)[colnames(FF) %in% myContinuous]
myBinary<-FFmyMeta%>%filter(type=="Binary")%>%pull(old_name); myBinary<-colnames(FF)[colnames(FF) %in% myBinary]
myCategorical_O<-FFmyMeta%>%filter(type=="Ordered Categorical")%>%pull(old_name); myCategorical_O<-colnames(FF)[colnames(FF) %in% myCategorical_O]
myCategorical_U<-FFmyMeta%>%filter(type=="Unordered Categorical")%>%pull(old_name); myCategorical_U<-colnames(FF)[colnames(FF) %in% myCategorical_U]
#filter to only vars of interest - not continuous variables  
myRecodes<-recodes_long %>% filter(old_name %in% c(myBinary, myCategorical_O, myCategorical_U))
#apply matchmaker
FF_labeled<-match_df(FF, dictionary=myRecodes, from = 'old_values', to='new_labels', by='old_name')
#reapply correct variable class types and na_codes
na_codes<-c("-1 Refuse", "-2 Don't know", "-3 Missing", "-9 Not in wave")
FF_labeled<-FF_labeled %>% mutate_at(vars(any_of(c(myBinary, myCategorical_O, myCategorical_U))), funs(as.factor(ifelse(. %in% na_codes, "Missing", .)))) %>% 
          mutate_at(vars(any_of(myContinuous)), funs(as.numeric(na_if(na_if(na_if(na_if(., "-3"), '-9'), '-6'), '-2'))))
#read back in the variable labels 
FF_labeled<-set_label(FF_labeled, get_label(FF))
FF_labels<-FF_labeled
```

## Cleaning pdqc data 
### Figure out analytic subsets 1) any methylation data; 2) 2-visits of methylation data
```{r, pdqcReadin}
pdqc_all<-readRDS(paste0(datadir, 'OGData/', "pd_qc.rds"))
pdqc<-pdqc_all %>% filter(probe_fail_10==0 & sex==predicted_sex)
#What FF ids have methylation data?
methy_ids<-pdqc$id
#create slide variable fo pdqc 
pdqc$slide <- gsub('_.*', '', pdqc$MethID)
pdqc$array<-gsub('.*_', '', pdqc$MethID)
#ids with 9visit, 15visit or both visit
visit9ids<-pdqc%>%filter(childteen=='C')%>%pull(idnum)%>%unique()
visit15ids<-pdqc%>%filter(childteen=='T')%>%pull(idnum)%>%unique()
bothvisit_ids<-visit9ids[visit9ids %in% visit15ids]
visit9only_ids<-visit9ids[!(visit9ids %in%bothvisit_ids)]
visit15only_ids<-visit15ids[!(visit15ids %in%bothvisit_ids)]
```

```{r, createAnalyticSubsetFlags}
#create flag in FF for having methylation data
#and flag for having 2 visits of methylation data
FF_labeled<-FF_labeled %>% 
  mutate(MethylData=ifelse(id %in% methy_ids, "Analysis subset", "Not in analysis subset"),
         TwoVisitData=ifelse(id%in%bothvisit_ids,"Analysis subset - both visits", "Not in both visit analysis subset"))
table(FF_labeled$MethylData)
table(FF_labeled$TwoVisitData)
```

```{r, checkIDlengths}
if((length(visit15only_ids)+length(visit9only_ids)+length(bothvisit_ids)!=length(FF_labeled %>% filter(MethylData=="Analysis subset")%>%pull(id)))){print("WARNING ID LENGTHS DO NOT ADD UP")}
if(length(bothvisit_ids)!=length(FF_labeled %>% filter(TwoVisitData=="Analysis subset - both visits")%>%pull(id))){print("WARNING ID LENGTHS DO NOT ADD UP")}
```

### Filter technical replicates (only 1 technical replicate per person-visit + decide based on QC variables - pick the one with the lower probe fail percentage)

```{r}
#technical replicates will be duplicate id + childteen variables from pdqc
pdqc<-pdqc %>% mutate(newID=paste(idnum, childteen, sep='_'))
techrep_ids<- pdqc %>% select(newID) %>% group_by(newID)%>%filter(n()>1)
techreps<-pdqc %>% filter(newID %in% techrep_ids$newID)%>%arrange(idnum,childteen)
hiPctPF<-techreps %>% group_by(newID) %>%filter(probe_fail_pct==max(probe_fail_pct))%>%pull(MethID)
pdqc_clean<-pdqc %>% filter(!(MethID %in% hiPctPF) & childteen!='M')
myMethIDs<-pdqc_clean %>% pull(MethID)
```

## Methylation summary variables

### Cell type proportions

Cell type proportions estimated using [epidDISH](https://www.bioconductor.org/packages/release/bioc/vignettes/EpiDISH/inst/doc/EpiDISH.html#how-to-estimte-cell-type-fractions-using-dnam-data) package and the EpiFibIC reference (generic for epithelial tissues).

```{r read in betaqc & celltype proportions}
betaqc<-readRDS(file=paste0(datadir, 'OGData/', "noob_filtered.rds"))
aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
globalmethy<-colMeans(betaqc, na.rm=T)
globalmethdf<-as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy))
aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
aprioriCGdf$MethID = colnames(betaqc)
#cell type proportions 
celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
estF_FF<-as.data.frame(celltypes$estF)
estF_FF$MethID = row.names(estF_FF)
```

### Clocks

```{r, readInclocks}
clocks<-read.table(file=paste0(datadir, "OGData/DNAm_clocks_ffcw.txt"), header = T)
colnames(clocks)[1]<-"MethID"
```

### Global methylation score

```{r make polymethylaton score}
#read in cpgs from PMC4833289
cpgs<-read.csv(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
#make headers
header1<-scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
header2<-paste(header1, scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
names(cpgs)<-header2
#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
#pivot_longer to get single column of coefs and a column for the analysis type
#and group into a list by analysis type
cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%group_split(set)
cpgs<-cpgs%>%setNames(lapply(cpgs, function(x){x$set %>% unique()}))
#make scores - no transform, zscore and centermean
#first, pull script 
source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
scores=makePolyEpiScore(m=betaqc, b=cpgs)
scores_z=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'zscore')
scores_center=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'meancenter')
polyscores=list('notransform'=scores, 'zscore'=scores_z, 'center'=scores_center)
polyscores<-lapply(polyscores, function(x) x%>% mutate(MethID=colnames(betaqc)))
save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores.RDS'))
polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% setNames(c(paste0(colnames(polyscores[[i]])[1:4], '_', names(polyscores)[i]), 'MethID')))%>%reduce(left_join, by='MethID')
```

```{r make methyl dataframe}
clocks2<-clocks %>% mutate(idnum=gsub('a', '', idnum))%>%filter(MethID %in% pdqc$MethID) # currently clocks are missing 3 individuals who were in johns pipeline but not jonahs

methyldata<-left_join(left_join(left_join(left_join(left_join(pdqc_clean %>% select(MethID, slide, array,childteen, idnum), globalmethdf), clocks2), estF_FF), aprioriCGdf), polyscores_wide)
methyldata[, c("globalmethylation", aprioriCG)]<-lapply(methyldata[, c("globalmethylation", aprioriCG)],function(x) as.numeric(as.character(x)))
#lapply(methyldata, class)
methyldata_wide <- methyldata %>% select(-MethID)%>% pivot_wider(., names_from=childteen, values_from=any_of(colnames(methyldata)[4:length(colnames(methyldata))]))
```

```{r individual who is missing global methylation data but had apriori CpG and PMS data?, eval=F}
#weirdID<-methyldata %>% filter(is.na(globalmethylation))%>%pull(MethID)
#globalmethy[weirdID] #not actually missing global methylation data
#clocks %>% filter(MethID%in%weirdID) # #idnum from  not the same as in
#pdqc_clean %>% filter(MethID==weirdID) #this is just a typo in the idnum from data entry, see format of other #idnums
#pdqc_clean %>% pull(idnum) %>% head()
#solution: when creating methyldata, first trim alphabet from clocks$idnum that contain alphabetic characters --> implemented
#also misssing clocks for 3 sex outlier samples that weren't in jonahs' pipeline: 
table(is.na(methyldata$horvath13))
weirdIDs<-methyldata %>% filter(is.na(horvath13))%>%pull(MethID)
pdqc_all %>% filter(MethID %in% weirdID) %>% xtabs(~predicted_sex_outlier, data=.)
```

## Creation of new variables
### Binary smoking variable
```{r}
FF_labeled<-FF_labeled %>%mutate(smkPreg_binary=ifelse(m1g4%in% c("1 2+pk/d","2 1<pk<2","3 <1pk/d" ), "Yes", ifelse(m1g4%in%c("-2 Don't know", "-3 Missing", "-9 Not in wave"), NA, ifelse(m1g4== "4 None", "No", ifelse(m1g4=="Missing", "Missing", "FLAG")))))
#xtabs(~m1g4+smkPreg_binary, data=FF_labeled)
```

### Ancestry variables

Ancestry is based on csv files of ancestry-specific PCS created by Erin (per Jonah meeting Oct 6 2020). If an ID was in {african/european/hispanic}, then the individual is labeled as such. 
```{r}
# read in PCs
euroPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/european.csv")
afrPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/african.csv")
hisPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/hispanic.csv")
allPC<-read.table("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/global_pcs.txt", col.names=c("X", "idnum", paste("PC", 1:20, sep='')))
#individuals in each ancestry specific pc group can be labeled as that ancestry for categorical ancestry variable 
FF_labeled<-FF_labeled %>% mutate(ancestry=case_when(MethylData=="Analysis subset" & id %in% hisPC$idnum~"Hispanic ancestry", MethylData=="Analysis subset" & id %in%afrPC$idnum~"African ancestry", MethylData=="Analysis subset" & id %in% euroPC$idnum~"European ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & (id %in% allPC$idnum)~"Other ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & !(id %in% allPC$idnum)~"Missing PC data", MethylData!="Analysis subset"~"Not in analysis subset"))
#check ancestry variable
#table(FF_labeled$ancestry)
#with(FF_labeled, xtabs(~ancestry+cm1ethrace))
MissingPCdata<-FF_labeled %>% filter(ancestry=="Missing PC data")%>%select(id)
save(MissingPCdata, file = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/CreatedData/MissingPC.Rdata")
#add PC data to FF_labeled 
FF_labeled<-left_join(FF_labeled, allPC %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>%rename_at(vars(-id), function(x) paste0('global_', x, sep='')))
localPC<-rbind(euroPC, afrPC, hisPC)
FF_labeled<-left_join(FF_labeled, localPC  %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>% rename_at(vars(-id), function(x) paste0('local_', x, sep='')))
```

There were `r length(MissingPCdata$id)` individuals without PC data. None of these individuals had data in the global PC file, so I assume they are actually missing genetic data (for whatever reason, possibly QC).

### Prenatal exposure variables 
Recoded frequency of prenatal alcohol drinking and drug use into Yes/no binary variables for drinking and drug use during pregnancy. 
```{r PrenatalExposure}
#table(FF_labeled$m1g2) # during preg, alchohol freq
#table(FF_labeled$m1g3) # during preg, drugs freq 
with(FF_labeled, table(m1g2, m1g3))
FF_labeled<-FF_labeled %>% mutate_at(c("m1g2", "m1g3"), .funs=funs(YesNoPreg=ifelse(.=="5 Never", "No", ifelse(. =="Missing", "Missing", "Yes"))))
with(FF_labeled, table(m1g2, m1g2_YesNoPreg))
with(FF_labeled, table(m1g3, m1g3_YesNoPreg))
#table(FF_labeled$m1g5) # in last year, alc/drugs interfered with work/relationships
#table(FF_labeled$m1g6) # ever sought help for drug/alc problems
```
## Postnatal smoke exposure 
Maternal postnatal smoking at different survey points is fairly correlated, and alluvial plots of changes over time show broad consistency, so I decided to create a single smoking variable for maternal smoking at age 1 and age 5 - if mothers were smoking at either age 1 or age 5 then "Maternal smoking at age 1 or 5", if not smoking at age 1 and age 5 then "No maternal smoking at age 1 and age 5" and then if missing data at one age and no at the other, then "Missing" (and if missing at both ages, "Missing"). Also created a similar dose variable capturing maternal packs per day that could be used instead. Decided to use just age 1 and age 5 because at these two time points the mother was asked (as opposed to PCG, also the dataset of Fragile Families metadata is missing PCG questions from age 3 about smoking) and they both preceed the actual saliva collection for both saliva collection age points. Then use a smoking dose variable from the mother or PCG interview at wave of saliva collection as an additional second hand smoke exposure variable.

```{r PostnatalSmoke}
#Baseline (1) -> Year 1/Age 1 (2) ->Year 3/Age3 (3) -> Year 5 (4) ->Year 9 (5) -> year 15 (6)
#Do you smoke in past month (how many packs per day)
#Mom            ->m2j5 (m2j5a)    ->                  -> m4j18 (m4j19) /p4 -> m5g17 (m5g18) / n5f17 (n5f18) 
#PCG                              p3a23a(p3a23b)                              n5f17 (n5f18)-> p6h76 (p6h77)
#seem to be missing p3 variables (PCG survey year 3) 
#number of people in household who smoke 
#p3a23->p5h15b->p6h78

#hours child around smoke 
#p3a21 -> p4a24->p5h15c

#with(FF_labeled, table(m2j5, m4j18))
#with(FF_labeled, table(m4j18, m5g17))
#with(FF_labeled, table(n5f17, m5g17))
#with(FF_labeled, table(p6h76, m5g17))

plot_grid(FF_labeled %>% select(m2j5, m4j18, m5g17, n5f17, p6h74) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between Yes/No smoking (mother) \n at years 1, 5, & 9 & PCG at years 9 & 15")
,FF_labeled %>% select(m2j5a, m4j19, m5g18, n5f18, p6h77) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between packs per day smoking (mother) \n at years 1, 5, 9 & PCG at years 9 and 15"), nrow=2)

yesnosmkalluv<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5, m4j18, m5g17, p6h74) %>% mutate(m5g17=stringr::str_to_title(m5g17))%>% group_by(m2j5, m4j18, m5g17, p6h74) %>% count()), axes=1:4, id='Cohort')

#ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow(stat="alluvium", lode.guidance="backfront", color='darkgrey')+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))

smkdose<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5a, m4j19, m5g18, p6h77)%>% mutate_at(vars(c("m2j5a", "m4j19", "m5g18")), funs(factor(substring(., 1, 1), labels=levels(FF_labeled$m2j5a)))) %>% group_by(m2j5a, m4j19, m5g18, p6h77) %>% count()), axes=1:4, id='Cohort')
#ggplot(smkdose, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))

plot_grid(ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))
+ggtitle("Alluvial plot of Yes/No smoking variables from age 1 to age 9 (mother) and age 15 (PCG)"), ggplot(smkdose %>% filter(!(stratum %in% c("-6 Skip", "Missing"))), aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))+ggtitle("Alluvial plot of smoking per day among those smoking from age 1 to age 9 (mother) and 15 (PCG)"), nrow=2)

#ggplot(FF_labeled, aes(x=p5h15b, y=p6h78))+geom_jitter()+stat_cor()+scale_x_continuous('# smoking in house age 9')+scale_y_continuous('# smoking in house age 15')+ggtitle("# individuals smoking in house correlation age 9 and age 15")

#Make any maternal postnatal smoking variable & dose categorization for maternal smoking at ages 1 & 5 
FF_labeled<-FF_labeled %>% mutate(PostnatalMaternalSmokingAny = case_when(
                                m2j5 =="1 Yes" ~ "Maternal smoking at age 1 or age 5",
                                m4j18=="1 Yes" ~ "Maternal smoking at age 1 or age 5",
                                m2j5 == "Missing" & m4j18=="Missing" ~ "Missing",
                                m2j5 == "Missing" & m4j18=="2 No" ~ "Missing",
                                m2j5 == "2 No" & m4j18=="Missing" ~ "Missing",
                                m2j5 == "2 No" & m4j18=="2 No" ~ "No maternal smoking at age 1 and age 5")) %>%
  mutate(PostnatalMaternalSmokingDose=case_when(
                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5   
                                          More 2 packs/day") ~ "Mom - pack/d or greater at both age 1 & 5", 
                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
                                !(m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
                                          More 2 packs/day")) ~ "Mom - pack/d or greater at either age 1 or 5",
                                !(m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d")) &
                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
                                          More 2 packs/day") ~ "Mom - pack/d or greater at either age 1 or 5",
                                m2j5a %in% c("1 <1/2 pk/d") | m4j19 %in% c("1 Less half pack/day")~"Mom - less than half pack/d at either age 1 or 5", 
                                m2j5a %in% c("-6 Skip") & m4j19 %in% c("-6 Skip")~
                                  "No maternal smoking at age 1 or 5", 
                                m2j5a %in% c("Missing") & m4j19 %in% c("-6 Skip", "Missing")~"Missing", 
                                m2j5a %in% c("Missing", "-6 Skip") & m4j19 %in% c("Missing")~"Missing"))

with(FF_labeled, table(PostnatalMaternalSmokingAny, m2j5, m4j18))
#with(FF_labeled, table(m2j5a, m4j19, PostnatalMaternalSmokingDose))
#with(FF_labeled, table(PostnatalMaternalSmokingAny, PostnatalMaternalSmokingDose))
```

Maybe here best to have 1 variable capturing any maternal postnatal smoking at age 1 & 5 and then use in each crossectional model the dose of maternal postnatal smoking at current visit (and similarly for linear mixed effect model)?

```{r makeMyFF}
varLabels2<-c(paste(colnames(FF_labeled)[1:66], varLabels), "Has any methylation data?", "Has two visits of methylation data", "Any maternal smoking during pregnancy", "Ancestry from PCA of child's genetic data", "PC 1 (global)", "PC 2 (global", "PC 1 (within ancestry category)", "PC 2 (within ancestry category",
"Any alcohol during pregnancy", "Any drugs during pregnancy", "Any postnatal smoking (age 1 & 5)", "Postnatal smoking dose at age 1 & 5")
FF_labeled<-set_label(FF_labeled, varLabels2)

myFF<-FF_labeled %>% filter(MethylData=="Analysis subset")%>%mutate(idnum=id)
myFF<-left_join(myFF, methyldata)
varLabels3<-c(varLabels2, "ID", "Methylation ID", "Slide", "Array (RC)", "Visit", "Global methylation", "horvath13 methylation clock", "horvath18 methylation clock", "pediatric methylation clock", "levine methylation clock", "Epithelial cell proportion", "Fibroid cell proportion", "Immune cell proportion", "cg05575291 methylation", "cg04180046 methylation",  "cg05549655 methylation", "cg14179389 methylation", "PES: Any smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood (+CT control)", "PES: Sustained smoking, older children peripheral blood", "PES: Any smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood (+CT control) - zscore", "PES: Sustained smoking, older children peripheral blood - zscore", "PES: Any smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood (+CT control) - mean center", "PES: Sustained smoking, older children peripheral blood - mean center")
myFF<-set_label(myFF, varLabels3)
```
### Child smoking
Exclude children who reported smoking at age 9 and age 15. Interestingly, the children who report smoking at age 9 do not report smoking at age 15, and the parental perception of child smoking at age 9 does not match the children who say they have ever smoked a cigarrette. For now using the 'ever smoked' variable from age 15 for the exclusion, but there is also a 'smoked in past month' that we could use instead (lose fewer children).

```{r, ChildSmoke}
with(myFF, table(k5f1l,childteen)) # 6 kids smoking at age 9 
with(myFF, table(p5q3cr, k5f1l)) # not the same kids whose parents say somewhat or very true that the child is using tobacco
with(myFF, table(k6d40, childteen)) # 41 kids say ever smoked a cigarette at age 15
#with(myFF, summary(k6d41))
with(myFF, table(k6d42, childteen)) # of these only 13 say they've smoked in the past month 
#with(myFF, summary(k6d43))
with(myFF, table(k5f1l, k6d40, childteen))# not the same kids at age 9 and 15
```

### Maternal or PCG smoking dose at time of sample

```{r visitMaternalSmoke}
myFF<-myFF %>% mutate(SmkAtVisitPastmonth=case_when(childteen=='C' & m5g17=='1 yes' & m5g18 %in% c("2 about a pack", "3 a pack and a half", "4 about 2 packs", 
                              "5 more than two  packs") ~ "Pack or more a day", 
                              childteen=='C' & m5g17=='1 yes' & m5g18=="1 less than half a pack a day" ~ "Less than pack a day", 
                              childteen=='C' & n5f17=="1 yes" & n5f18 =='2 about a pack' ~ "Pack or more a day",
                              childteen=='C' & n5f17=="1 yes" & n5f18 =="1 less than half a pack day" ~ "Less than pack a day",
                              childteen=='C' & m5g17=="1 yes" & m5g18 %in% c("-6 Skip", "Missing")~"Missing",
                              childteen == 'C' & n5f17%in%c("-7 N/A", "2 no") & m5g17=="2 no" ~ "No smoking", 
                              childteen == 'C' & n5f17%in%c("2 no") & m5g17%in%c("2 no", "Missing") ~ "No smoking", 
                              childteen == 'C' & n5f17 %in% c("Missing", "-7 N/A") & m5g17 =="Missing" ~ "Missing", 
                              childteen=='T' & p6h76 %in% c("-6 Skip", "0 None") ~ "No smoking", 
                              childteen=='T' & p6h77%in% c("1 5 or fewer cigarettes per day", "2 About half a pack a day, 10 cigarettes")~"Less than pack a day", 
                              childteen=='T' & p6h77 %in% c("3 About a pack a day, 20 cigarettes", "4 More than 1 pack a day")~"Pack or more a day",
                              childteen=='T' & p6h77 == "Missing"~"Missing"))%>%
              mutate(SmkAtVisitYesNo=case_when(SmkAtVisitPastmonth=="No smoking"~"No", 
                                               SmkAtVisitPastmonth %in% c("Less than pack a day", "Pack or more a day") ~ "Yes", 
                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & !(p6h76 %in% c("-6 Skip", "0 None", "Missing"))~"Yes",
                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17=='1 yes'~"Yes", 
                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17!='1 yes'~"Missing", 
                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & p6h76 %in% c("-6 Skip", "0 None", "Missing")~"Missing"))

myFF %>% filter(childteen=='C') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
myFF %>% filter(childteen=='T') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
myFF$SmkAtVisitPastmonth<-factor(myFF$SmkAtVisitPastmonth)
```
### Child's age at time of interview/visit
```{r}
myFF %>% filter(is.na(cp6yagem) &is.na(hv6_yagem) & childteen=='T')%>%dim() #3
myFF %>% filter(is.na(cm5b_age) &is.na(hv5_agem) &childteen=='C')%>%dim() #1 
myFF %>% filter(!is.na(cm5b_age) & is.na(hv5_agem) & childteen=='C')%>%dim()#1
myFF %>% filter(!is.na(cp6yagem) &is.na(hv6_yagem))%>%dim() #990

myFF <- myFF %>% 
  mutate(ChildAgeAtIn=case_when(childteen=='C'~ cm5b_age, childteen=='T'~cp6yagem))%>%
  mutate(ChildAgeAtVisit=case_when(childteen=='C'~hv5_agem, childteen=='T'~hv6_yagem))%>% 
  mutate(ChildAgeComposite=case_when(childteen=='C' & !is.na(hv5_agem)~hv5_agem, 
         childteen=='C' & is.na(hv5_agem)~cm5b_age, 
         childteen=='T' & !is.na(hv6_yagem)~hv6_yagem, 
         childteen=='T' & is.na(hv6_yagem)~cp6yagem))

myFF %>% filter(childteen=='C')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)

myFF %>% filter(childteen=='T')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)

ggplot(myFF%>%filter(childteen=='C'), aes(cm5b_age, hv5_agem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (9 yr) and childs age at home visit (9 yr)")
age5resid<-myFF %>% filter(childteen=='C')%>%mutate(age5_resid=cm5b_age-hv5_agem)%>%pull(age5_resid)
hist(age5resid)

>>>>

# Objectives & Hypotheses 
 A) Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen. B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking. C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and support vector machines and evaluate the agreement of the classifier at two time points.

Hypotheses: I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups. I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time. Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.

@@

## Labeling phenotype data
```{r, phenotypedataReadinLabel, include=F}
pheno<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/fullpheno.rds")
